---
title: "Review Summary"
format: html
execute:
  echo: false
editor: visual
---

```{r}
#| label: setup
#| include: false

library(googlesheets4)
library(tidyverse)

```

```{r}
#| label: read-data
#| include: false

gs4_deauth() # Just so Google doesn't try to authenticate you}
read_review <- function(sheet) {
  read_sheet(
    "https://docs.google.com/spreadsheets/d/1zRA77pWp4nFZ9sjhVKXxxe0TBSg96QBC0L8XThshfQE/edit?usp=sharing",
    sheet = sheet,
    skip = 1,
    na = c("", "NA")
  )
}
papers <- read_review("papers")
devices <- read_review("devices")
taxa <- read_review("taxa")
datasources <- read_review("data")

```

```{r}
#| label: clean-data

novel_data_papers <- papers %>% 
  drop_na(manuscript_type) %>% 
  filter(manuscript_type != "U",
         novel_biologging == "Y",
         biologging_context == "W")
novel_data_devices <- devices %>% 
  select(-note) %>% 
  semi_join(novel_data_papers, by = "id")
novel_data_taxa <- devices %>% 
  select(-note) %>% 
  semi_join(novel_data_papers, by = "id")
novel_data_sources <- datasources %>% 
  select(-note) %>% 
  semi_join(novel_data_papers, by = "id")

```

## When and where

`r nrow(novel_data_papers)` papers presented novel bio-logging data. The number of papers published per year is trending up (max of 12 in 2021).

```{r}
#| label: paper-years

novel_data_papers %>% 
  count(year) %>% 
  complete(year = min(year):max(year), fill = list(n = 0)) %>% 
  ggplot(aes(year, n)) +
  geom_line() +
  geom_point() +
  theme_minimal()

```

*Marine Ecology Progress Series* published the most papers of any journal, followed by *Animal Biotelemetry* and *PLOS One*. Only journals with two or more publications shown here.

```{r}
#| label: journals

fmt_source <- function(source) {
  wrap_lbl <- function(lbl) {
    lbl %>% 
      strwrap(20) %>% 
      paste(collapse = "\n")
  }
  source %>% 
    fct_infreq() %>% 
    fct_relabel(\(lbl) map_chr(lbl, wrap_lbl))
}

novel_data_papers %>% 
  mutate(source = fmt_source(source)) %>% 
  count(source) %>% 
  filter(n > 1) %>% 
  ggplot(aes(source, n)) +
  geom_col() +
  expand_limits(y = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 60),
        axis.title.x = element_blank())
```

Location sensors (GPS, Argos, and depth) are used in the most studies. Accelerometers and temperature are the most commonly used intrinsic and environmental sensors, respectively.

```{r}
#| label: device-types

novel_data_devices %>% 
  pivot_longer(c(ltype, itype, etype), 
               names_to = "cat", 
               values_to = "sensor") %>% 
  separate_longer_delim(sensor, ",") %>% 
  mutate(cat = factor(cat,
                      levels = c("ltype", "itype", "etype"),
                      labels = c("Location", "Intrinsic", "Environment")),
         sensor = str_trim(sensor, "both")) %>% 
  filter(sensor != "?") %>% 
  drop_na(sensor) %>% 
  count(cat, sensor) %>% 
  mutate(sensor = fct_reorder(sensor, n, .desc = TRUE)) %>% 
  ggplot(aes(sensor, n, fill = cat)) +
  geom_col() +
  scale_fill_brewer("Sensor type", palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 60))

```
