---
title: "Power analysis"
format: html
editor: visual
---

```{r}
#| label: setup

library(brms)
library(cowplot)
library(HDInterval)
library(tidybayes)
library(tidyverse)
set.seed(1234)

inv_logit <- function(x) exp(x) / (1 + exp(x))
# Find the minimum Highest Density Interval in draws that contains value
find_min_hdi <- function(draws, value, tol = 1e-3) {
  lwr <- tol
  upr <- 1 - tol
  cred <- mean(c(lwr, upr))
  int <- hdi(draws, cred)
  while (upr - lwr > tol) {
    if (between(value, int[1], int[2])) {
      upr <- cred
    } else {
      lwr <- cred
    }
    cred <- mean(c(lwr, upr))
    int <- hdi(draws, cred)
  }
  cred
}

```

## Simulated data

Simulation specified by:

$$
\begin{align}
DV_i &\sim Bernoulli(p_i) \\
logit(p_i) &= \alpha + \beta_e[e_i] + \beta_s[s_i] + \beta_t \times t_i
\end{align}
$$

Where:

-   $DV$ is a dependent variable ( $f$, $a$, or $r$ )

-   $p$ is the probability of the dependent variable

-   $\alpha$ is the intercept

-   $\beta_e$ and $\beta_s$ are coefficients for the categorical predictors *ecosystem* and *sensor class*, and $e$ and $s$ are the ecosystem and sensor class of the observation

-   $\beta_t$ is the coefficient for time, $t_i$, which is quantified as "years since 2007"

For this analysis, assume the following values for the coefficients:

$$
\begin{align}
\alpha &= -4 \\
\beta_e &= [marine = 0, terrestrial = 0.5] \\
\beta_s &= [aspatial = 0, spatial = 1.5] \\
\beta_t &= 0.2
\end{align}
$$

The independent variables $t$, $e$, and $s$ are generated through stratified sampling. We simulate equal sample sizes from all combinations of the following categories:

-   $e$ in *marine*, *terrestrial*

-   $s$ in *aspatial*, *spatial*

-   Early (2007 - 2015), middle (2016-2019), and recent (2020-2023) periods

$t$ will be uniformly distributed within each period.

For the power analysis, we vary the sample size ( $n^*$ ) within each of the 12 combinations of categories. For example, if $n^*=50$, then total sample size $n=12 \times n^*=600$. Simulated data for $n^*=50$ shown in figure [@fig-n50].

```{r}
#| label: sim-params

alpha <- -4
beta_E <- c(marine = 0, terrestrial = 0.5)
beta_S <- c(aspatial = 0, spatial = 1.5)
beta_T = 0.2

```

```{r}
#| label: fig-n50
#| fig-caption: "Simulated data for dependent variable $DV$ when sampling 50 papers within each combination of categories for $e$, $s$, and $t$. Points represent simulated studies, lines represent probability of the dependent variable $p(DV)$. Gold indicates terrestrial, navy indicates marine. Triangle points/dashed lines represent spatial data, circle points/solid lines represent aspatial data."

biolog_n50 <- expand_grid(
    era = c("early", "middle", "recent"),
    E = c("marine", "terrestrial"),
    S = c("aspatial", "spatial")
  ) %>%
    slice(rep(seq(nrow(.)), each = 50)) %>%
    mutate(.id = seq(nrow(.)),
           T = case_when(
             era == "early"  ~ sample(2007:2015, nrow(.), replace = TRUE),
             era == "middle" ~ sample(2016:2019, nrow(.), replace = TRUE),
             era == "recent" ~ sample(2020:2023, nrow(.), replace = TRUE)
           )) %>%
    mutate(
      T2007 = T - 2007,
      logit_p = alpha + beta_E[E] + beta_S[S] + beta_T * T2007,
      p = inv_logit(logit_p),
      DV = rbinom(nrow(.), size = 1, prob = p)
    )

e_palette <- c(marine = "navy", terrestrial = "goldenrod")

biolog_text <- filter(biolog_n50, T == 2023) %>%
  distinct(E, S, p) %>%
  mutate(label = str_to_title(paste(E, S, sep = "\n")))

ggplot(biolog_n50, aes(T, p, color = E)) +
  geom_jitter(aes(y = DV, shape = S),
              width = 0.25, height = 0.05,
              alpha = 0.5) +
  geom_line(aes(linetype = S)) +
  geom_text(aes(label = label), biolog_text,
            x = 2023.1, hjust = 0) +
  scale_color_manual(values = e_palette) +
  scale_x_continuous(limits = c(NA, 2025),
                     breaks = seq(2010, 2020, by = 5)) +
  labs(y = expression(italic(p(DV)))) +
  expand_limits(x = 2025) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title.x = element_blank())

```

We use the following priors:

$$
\begin{align}
\alpha &\sim Normal(0, 4) \\
\beta_{e=terrestrial} &\sim Normal(0, 2) \\
\beta_{s=spatial} &\sim Normal(0, 2) \\
\beta_t &\sim Normal(0, 0.5)
\end{align}
$$

## Power analysis

How well does the model recover the simulation parameters with varying sample sizes? We use the *smallest highest density interval* of the linear predictor ( $logit(p)$ ) that contains the actual values as our metric. This metric combines the uncertainty, We simulate data 100 times each for $n^*$ values of 10:10:100. Then, we fit a model to the simulated data and .

```{r}
#| label: power-analysis

sim_analysis <- function(nstar) {
  biolog <- expand_grid(
    era = c("early", "middle", "recent"),
    E = c("marine", "terrestrial"),
    S = c("aspatial", "spatial")
  ) %>%
    slice(rep(seq(nrow(.)), each = nstar)) %>%
    mutate(
      T = case_when(
        era == "early"  ~ sample(2007:2015, nrow(.), replace = TRUE),
        era == "middle" ~ sample(2016:2019, nrow(.), replace = TRUE),
        era == "recent" ~ sample(2020:2023, nrow(.), replace = TRUE)
      ),
      T2007 = T - 2007,
      logit_p = alpha + beta_E[E] + beta_S[S] + beta_T * T2007,
      p = inv_logit(logit_p),
      DV = rbinom(nrow(.), 1, p)
    )
  biolog_prior <- c(
    set_prior(prior = "normal(0, 4)", class = "Intercept"),
    set_prior(prior = "normal(0, 2)", coef = "Eterrestrial"),
    set_prior(prior = "normal(0, 2)", coef = "Sspatial"),
    set_prior(prior = "normal(0, 0.5)", coef = "T2007")
  )
  biolog_mod <- brm(
    DV ~ E + S + T2007,
    data = biolog,
    family = bernoulli(link = "logit"),
    prior = biolog_prior,
    save_pars = save_pars(all = TRUE),
    chains = 4,
    cores = 4,
    iter = 50000,
    refresh = 0
  )
  biolog_actual <- expand_grid(T2007 = (2007:2023) - 2007, 
                               E = c("marine", "terrestrial"), 
                               S = c("aspatial", "spatial")) %>% 
    mutate(logit_p = alpha + beta_E[E] + beta_S[S] + beta_T * T2007)
  biolog_draws <- biolog_mod %>% 
    as_draws_df() %>% 
    as_tibble() %>% 
    cross_join(biolog_actual) %>% 
    mutate(linpred_post = b_Intercept + 
             b_Eterrestrial * (E == "terrestrial") + 
             b_Sspatial * (S == "spatial") + 
             b_T2007 * T2007)
  
  min_hdi <- biolog_draws %>% 
    group_by(E, S, T2007) %>% 
    summarize(min_hdi = find_min_hdi(linpred_post, logit_p[1]),
              .groups = "drop") %>% 
    pull(min_hdi) %>% 
    max()
  hdi <- HDInterval::hdi(biolog_draws$linpred_post, min_hdi)
  viz <- biolog_draws %>% 
    group_by(T2007, E, S) %>% 
    summarize(
      across(linpred_post, 
             list(mean = mean, 
                  lwr = \(x) hdi(linpred_post, min_hdi)[1], 
                  upr = \(x) hdi(linpred_post, min_hdi)[2])), 
      .groups = "drop") %>% 
    ggplot(aes(T2007, color = E, fill = E, linetype = S)) + 
    geom_ribbon(aes(ymin = linpred_post_lwr, ymax = linpred_post_upr), 
                color = NA, alpha = 0.5) + 
    geom_line(aes(y = linpred_post_mean)) + 
    scale_color_manual(values = e_palette) + 
    scale_fill_manual(values = e_palette) + 
    geom_line(aes(y = logit_p), biolog_actual, linewidth = 1.5) + 
    labs(title = sprintf("N* = %d", nstar),
         caption = sprintf("HDI = %0.3f", min_hdi)) +
    theme_classic()
  list(min_hdi = min_hdi,
       viz = viz)
}

nstar <- seq(10, 100, by = 30)
niter <- 2
power_analysis <- map(
  nstar, 
  \(n) map(seq(niter), \(i) sim_analysis(n))
)
names(power_analysis) <- paste0("nstar=", nstar)

```

```{r}
#| label: power-analysis-summary
#| lightbox: true

map2(power_analysis, names(power_analysis), 
     \(hdi, n) tibble(nstar = n, hdi = hdi)) %>% 
  list_rbind() %>% 
  mutate(nstar = as.integer(substr(nstar, 7, 8))) %>% 
  ggplot(aes(nstar, hdi)) +
  geom_jitter(height = 0, width = 2) +
  expand_limits(y = c(0, 1)) +
  theme_classic()

```
